DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-21 manuscript/manuscript-reservoir.tex
Pequenas mudanças no texto

To a homogenous half-space, \citet{Mindlin&Cheng50} showed that the method proposed by  \citet{Goodier37} can be extended by the displacement solution given by:
\begin{equation}
\bf{u} = {\bf{\nabla}} \: \phi_{1} \: + \: {\bf{{\nabla}_{2}}} \phi_{2}, 
\label{eq:displacement-solution}
\end{equation}
where $\phi_{1} \equiv \phi_{1}(x,y,z)$ is the potential defined in equation \ref{eq:phi}, 
$\phi_{2} \equiv  \phi_{2}(x,y,z)$
is defined as "image potential" \citep{Segall92} due to a image point at the coordinates
$(x^{\prime}, y^{\prime}, -z^{\prime} )$ and the operator $\bf{{\nabla}_{2}}$ is expressed by
\begin{equation}
{\bf{{\nabla}_{2}}} = (3 - 4\nu){\bf{\nabla}} + 2 {\bf{\nabla}} z \frac{\partial }{\partial z}  - 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}_{z}}},
\label{eq:nabla2}
\end{equation}
where ${\bf{\hat{z}}}$ is the unit vector in the $z-$direction and 
${\bf{{\nabla}^{2}_{z}}}$ is a scalar operator in which the operand is first to be  multiplied by $z$ and then operated upon by the Laplacian ${\bf{\nabla}^{2}}$.

@@

By following \cite{Sharma56} and \cite{Tempone10}, 
the stress vector 
$\mbox{\boldmath$\sigma$}_{ij} \equiv \mbox{\boldmath$\sigma$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ 
at the point $(x_{i}, y_{i}, z_{i})$ due to the $j$th single nucleus of strain buried 
in the half space is given by
\begin{equation}
\mbox{\boldmath$\sigma$}_{ij} = \mbox{\boldmath$\sigma_{1}$}_{ij} + 
\mbox{\boldmath$\sigma_{2}$}_{ij} \: ,
\label{eq:stress_nucleus}
\end{equation}
where 
$\mbox{\boldmath$\sigma_{1}$}_{ij} \equiv \mbox{\boldmath$\sigma_{1}$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ is the stress vector 
at the point $(x_{i}, y_{i}, z_{i})$ due to the $j$th single nucleus in the infinite space and $\mbox{\boldmath$\sigma_{2}$}_{ij} \equiv \mbox{\boldmath$\sigma_{2}$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ is the stress vector 
at the point $(x_{i}, y_{i}, z_{i})$ that gives the correction of the stress due to the 
$j$th image nucleus considering a semi-space. These two vectors are given by
\begin{equation}
\mbox{\boldmath$\sigma_{1}$}_{ij} =
A  \: 
\begin{bmatrix} 
\frac{\partial^{2} }{\partial x \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
\: \Delta p_{j} \: \:dv_j^{\prime} \: ,
\label{eq:nucleus-stress1-vector}
\end{equation}
and 
\begin{equation}
\mbox{\boldmath$\sigma_{2}$}_{ij} = 
A  \: 
\left\{
\begin{bmatrix} 
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}} \\
- \frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}   
\end{bmatrix}
+ 2 \: z_{i}
\begin{bmatrix} 
\frac{\partial^{3}  }{\partial x \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial y \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial z^{3}} {\frac{1}{{R_2}_{ij}}} 
\end{bmatrix}
\right\}
\: \Delta p_{j} \: \:dv_j^{\prime} \: .
\label{eq:nucleus-stress2-vector}
\end{equation}
According to \cite{Sharma56} and \cite{Tempone10}, the Beltrami’s equations \citep{Beltrami} and the equilibrium equations must be satisfied to obtain the contribution of the stress field in the half space. 
Additionally, the boundary condition 
$\mbox{\boldmath$\sigma$}_{ij} = \mbox{ \boldmath$0$}$ (equation \ref{eq:stress_nucleus})
must be satisfied at the free surface ($z_i = 0$).
%\begin{equation}
%\mbox{\boldmath$\sigma$}_{ij} = \mbox{\boldmath$\sigma_{1}$}_{ij} + 
%\mbox{\boldmath$\sigma_{2}$}_{ij} = \mbox{ \boldmath$0$},
%\label{eq:null_stress}
%\end{equation}
%where $\mbox{ \boldmath$0$}$ is the null vector.
This condition can be easily verified by adding the vectors 
$\mbox{\boldmath$\sigma_{1}$}_{ij}$ (equation \ref{eq:nucleus-stress1-vector}) and 
$\mbox{\boldmath$\sigma_{2}$}_{ij}$ (equation \ref{eq:nucleus-stress2-vector}) 
computed at any point on the free surface $(x_i, y_i, z_i = 0)$.

@@

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $j$th prism
at the $i$th coordinates  $(x_i, y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:displacement_ui})
over its volume and obtain
\begin{equation}
{\bf{\tilde{u}}}_{ij} = 
\iiint\limits_{{v_{1}}_{j}}
{\bf{u_{1}}}_{ij} \: \:  dv_{j}^{\prime}
\: + \:
\iiint\limits_{{v_{2}}_{j}}
{\bf{u_{2}}}_{ij} \:\:  dv_{j}^{\prime} \: ,
\label{eq:nucleus-ui_jth_prism}
\end{equation}
where both integrals are conducted with respect to the variables 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. The first integral in equation 
\ref{eq:nucleus-ui_jth_prism} is conducted over the volume ${v_{1}}_{j}$ of the $j$th
prism and the second is conducted over the volume 
${v_{2}}_{j}$ of a different prism symmetrically positioned above the free surface 
and conveniently called $j$th "image prism".
Volume ${v_{1}}_{j}$ is defined by ${x_1}_{j}$, ${x_2}_{j}$, ${y_1}_{j}$, ${y_2}_{j}$,
${z_1}_{j}$, and ${z_2}_{j}$, which represent, respectively, the south, north, west, 
east, top, and bottom borders of the $j$th prism.
Volume ${v_{2}}_{j}$ of the $j$th image prism is defined in a similar way, but with 
top and bottom given by $z_1 - 2 \, z_c$ and $z_2 - 2 \, z_c$, where 
$z_c = \frac{1}{2}(z_1 + z_2)$ is the center depth of the $j$th prism.

@@

The remaining integrals depending on first and second derivatives 
of ${\frac{1}{{R_2}_{ij}}}$ in the right-hand side of equations 
\ref{eq:u_til_alpha} and \ref{eq:u_til_z} are conducted over the volume 
${v_{2}}_{j}$ of the $j$th image prism and have the following closed solutions 

@@

{Z_1}_{j} &= z_i - {z_1}_{j} + 2 \, z_c \\
{Z_2}_{j} &= z_i - {z_2}_{j} + 2 \, z_c

@@

where $z_c = \frac{1}{2}(z_1 + z_2)$ is the center depth of the $j$th prism. 

@@

In equations \ref{dx1}--\ref{dz1} and \ref{dx2}--\ref{dz1}, we adopted the modifications proposed by \cite{Fukushima2020}.
To overcome the zero division in evaluating the arguments of the arctangent function, \cite{Fukushima2020} replaced  $\tan^{-1} \big( \frac{S}{T} \big)$ by 
\begin{equation}
arctan2(S,T) = \begin{cases}
	%\left\{ \begin{aligned}
    atan (S/T) & \:\: \mbox{if} \: \:\:  T \neq 0 \\
    \pi /2 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S > 0 \\
    -\pi /2 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S < 0 \\
    0 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S = 0 \\
    %\end{aligned} \right\}.
    \end{cases}
\label{eq:arctan2}  
\end{equation}
If the argument of the logarithm is less than $10^{-10}$, the logarithm is replaced by zero; otherwise the logarithm is  calculated regularly

@@

Because we defined the $z-$axis as positive downwards, the positive vertical displacement means a subsidence and the negative vertical displacement means an uplift.
Figure \ref{fig:displacement}  shows the horizontal and vertical displacements  calculated, respectively, with equations \ref{eq:horizontal_displacement} and \ref{eq:u_til_z} by our methodology that uses the closed expressions of the full integrations (equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}) of \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1} $-$ \ref{dz2}).

@@

In both cases (Figures \ref{fig:displacement}b and \ref{fig:displacement_Geertsma}b) the vertical displacements due to the entire the disc-shaped reservoir display a subsidence (positive values) above the reservoir and an uplift (negative values) below the reservoir.
We stress that the proposed volume integrations  (equations \ref{dx1} $-$ \ref{dz2})  allowed  to evaluate the  vertical displacement (Figure \ref{fig:displacement}b) throughout  the entire reservoir including inside and outside the reservoir.
Rather, the  vertical displacement using Geertsma’s method 
(Figure \ref{fig:displacement_Geertsma}b) is only valid outside the reservoir.

@@

\codedataavailability{
The current version of our code is freely distributed under the BSD 3-clause licence and it is available for download at Zenodo: http://doi.org/XXXXXXXX. 
The latest development version of our code can be freely downloaded from a repository on GitHub (https://github.com/pinga-lab/reservoir-compaction). 
Instructions for running  the current version of our code are also provided on the repository.
The code is still being improved and we encourage the user to work with the latest development version. 
The code was developed as an open-source Python language (Python 3.7.6).
The numerical applications were produced in Jupyter Notebook. 
The data of the pore pressure distribution simulating a realistic reservoir 
$(realistic-model.pickle)$ are available  in the  above-mentioned repositories.
} %% use this section when having data sets and software code available

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement.png}
\caption{Reservoir under uniform depletion: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dz2})}
\label{fig:displacement}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_z_levels.png}
\caption{Reservoir under uniform depletion: (a) Horizontal x-component displacement and (b) vertical displacement by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dz2}).
These displacements are calculated along the x-axis, at $y = 0$ m and $z$ located at the depths of:  seafloor ($z = 0$ m), reservoir top ($z = 750$ m), reservoir center ($z = 800$ m) and reservoir bottom ($z = 850$ m).}
\label{fig:displacement_z_levels}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress.png}
\caption{Reservoir under uniform depletion: (a) $x-$, (b) $y-$, and (c) $z-$ components of the stress at the free surface.}
\label{fig:Null_stress}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dz2})}
\label{fig:displacement_non_uniform_depletion}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_z_levels_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) Horizontal x-component displacement and (b) vertical displacement by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dz2}).
These displacements are calculated along the x-axis, at $y = 0$ m and $z$ located at the depths of:  seafloor ($z = 0$ m), reservoir top ($z = 750$ m), reservoir center ($z = 800$ m) and reservoir bottom ($z = 850$ m).}
\label{fig:displacement_z_levels_non_uniform_depletion}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) $x-$, (b) $y-$, and (c) $z-$ components of the stress at the free surface.}
\label{fig:Null_stress_non_uniform_depletion}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_complex_reservoir.png}
\caption{Reservoir with arbitrary geometry and under arbitrary pressure changes: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dz2})}
\label{fig:displacement_complex_reservoir}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress_complex_reservoir.png}
\caption{Reservoir with arbitrary geometry and under arbitrary pressure changes: (a) $x-$, (b) $y-$, and (c) $z-$ components of the stress at the free surface.}
\label{fig:Null_stress_complex_reservoir}
\end{figure}

>>>>

To a homogenous half-space, \citet{Mindlin&Cheng50} showed that the method proposed by  \citet{Goodier37} can be extended by the displacement solution given by:
\begin{equation}
\bf{u} = {\bf{\nabla}} \: \phi_{1} \: + \: {\bf{{\nabla}_{2}}} \phi_{2}, 
\label{eq:displacement-solution}
\end{equation}
where $\phi_{1} \equiv \phi_{1}(x,y,z)$ is the potential defined in equation \ref{eq:phi}, 
$\phi_{2} \equiv  \phi_{2}(x,y,z)$
is defined as "image potential" \citep{Segall92} due to a image point at the coordinates
$(x^{\prime}, y^{\prime}, -z^{\prime} )$ and the operator $\bf{{\nabla}_{2}}$ is expressed by
\begin{equation}
{\bf{{\nabla}_{2}}} = (3 - 4\nu){\bf{\nabla}} + 2 {\bf{\nabla}} z \frac{\partial }{\partial z}  - 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}_{z}}},
\label{eq:nabla2}
\end{equation}
where ${\bf{\hat{z}}}$ is the unit vector in the $z-$direction and 
${\bf{{\nabla}^{2}_{z}}}$ is a scalar operator in which the operand is firstly
multiplied by $z$ and then operated upon by the Laplacian ${\bf{\nabla}^{2}}$.

@@

By following \cite{Sharma56} and \cite{Tempone10}, 
the stress vector 
$\mbox{\boldmath$\sigma$}_{ij} \equiv \mbox{\boldmath$\sigma$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ 
at the point $(x_{i}, y_{i}, z_{i})$ due to the $j$th single nucleus of strain buried 
in the half space is given by
\begin{equation}
\mbox{\boldmath$\sigma$}_{ij} = \mbox{\boldmath$\sigma_{1}$}_{ij} + 
\mbox{\boldmath$\sigma_{2}$}_{ij} \: ,
\label{eq:stress_nucleus}
\end{equation}
where 
$\mbox{\boldmath$\sigma_{1}$}_{ij} \equiv \mbox{\boldmath$\sigma_{1}$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ is the stress vector 
at the point $(x_{i}, y_{i}, z_{i})$ due to the $j$th single nucleus in the infinite space and $\mbox{\boldmath$\sigma_{2}$}_{ij} \equiv \mbox{\boldmath$\sigma_{2}$}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ is the stress vector 
at the point $(x_{i}, y_{i}, z_{i})$ that gives the correction of the stress due to the 
$j$th image nucleus considering a semi-space. These two vectors are given by
\begin{equation}
\mbox{\boldmath$\sigma_{1}$}_{ij} =
A  \: 
\begin{bmatrix} 
\frac{\partial^{2} }{\partial x \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
\: \Delta p_{j} \: \:dv_j^{\prime} \: ,
\label{eq:nucleus-stress1-vector}
\end{equation}
and 
\begin{equation}
\mbox{\boldmath$\sigma_{2}$}_{ij} = 
A  \: 
\left\{
\begin{bmatrix} 
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}} \\
- \frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}   
\end{bmatrix}
+ 2 \: z_{i}
\begin{bmatrix} 
\frac{\partial^{3}  }{\partial x \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial y \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial z^{3}} {\frac{1}{{R_2}_{ij}}} 
\end{bmatrix}
\right\}
\: \Delta p_{j} \: \:dv_j^{\prime} \: .
\label{eq:nucleus-stress2-vector}
\end{equation}
According to \cite{Sharma56} and \cite{Tempone10}, the Beltrami’s equations \citep{Beltrami} and the equilibrium equations must be satisfied to obtain the contribution of the stress field in the half space. 
Additionally, the boundary condition 
$\mbox{\boldmath$\sigma$}_{ij} = \mbox{ \boldmath$0$}$ (equation \ref{eq:stress_nucleus}), where $\mbox{ \boldmath$0$}$ is the zero vector, must be satisfied at the free surface ($z_i = 0$).
%\begin{equation}
%\mbox{\boldmath$\sigma$}_{ij} = \mbox{\boldmath$\sigma_{1}$}_{ij} + 
%\mbox{\boldmath$\sigma_{2}$}_{ij} = \mbox{ \boldmath$0$},
%\label{eq:null_stress}
%\end{equation}
%where $\mbox{ \boldmath$0$}$ is the null vector.
This condition can be easily verified by adding the vectors 
$\mbox{\boldmath$\sigma_{1}$}_{ij}$ (equation \ref{eq:nucleus-stress1-vector}) and 
$\mbox{\boldmath$\sigma_{2}$}_{ij}$ (equation \ref{eq:nucleus-stress2-vector}) 
computed at any point on the free surface $(x_i, y_i, z_i = 0)$.

@@

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $j$th prism
at the $i$th coordinates  $(x_i, y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:displacement_ui})
over its volume and obtain
\begin{equation}
{\bf{\tilde{u}}}_{ij} = 
\iiint\limits_{{v_{1}}_{j}}
{\bf{u_{1}}}_{ij} \: \:  dv_{j}^{\prime}
\: + \:
\iiint\limits_{{v_{2}}_{j}}
{\bf{u_{2}}}_{ij} \:\:  dv_{j}^{\prime} \: ,
\label{eq:nucleus-ui_jth_prism}
\end{equation}
where both integrals are conducted with respect to the variables 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. The first integral in equation 
\ref{eq:nucleus-ui_jth_prism} is conducted over the volume ${v_{1}}_{j}$ of the $j$th
prism and the second is conducted over the volume 
${v_{2}}_{j}$ of a different prism symmetrically positioned above the free surface 
and conveniently called $j$th "image prism".
Volume ${v_{1}}_{j}$ is defined by ${x_1}_{j}$, ${x_2}_{j}$, ${y_1}_{j}$, ${y_2}_{j}$,
${z_1}_{j}$, and ${z_2}_{j}$, which represent, respectively, the south, north, west, 
east, top, and bottom borders of the $j$th prism.
Volume ${v_{2}}_{j}$ of the $j$th image prism is defined in a similar way, but with 
top and bottom given by ${z_1}_{j} - 2 \, {z_c}_{j}$ and ${z_2}_{j} - 2 \, {z_c}_{j}$, where 
${z_c}_{j} = \frac{1}{2}({z_1}_{j} + {z_2}_{j})$ is the center depth of the $j$th prism.

@@

The remaining integrals, in the right-hand side of equations 
\ref{eq:u_til_alpha} and \ref{eq:u_til_z}, depend on first and second derivatives 
of ${\frac{1}{{R_2}_{ij}}}$.
These integrals are conducted over the volume ${v_{2}}_{j}$ of the $j$th image prism and have the following closed solutions \citep{Nagyetal2000, Nagyetal2002}:

@@

{Z_1}_{j} &= z_i - {z_1}_{j} + 2 \, {z_c}_{j} \\
{Z_2}_{j} &= z_i - {z_2}_{j} + 2 \, {z_c}_{j}

@@

where ${z_c}_{j} = \frac{1}{2}({z_1}_{j} + {z_2}_{j})$ is the center depth of the $j$th prism. 

@@

In equations \ref{dx1}--\ref{dz1}, \ref{dx2}--\ref{dzz2} and \ref{dxz1}--\ref{dzz1}, we adopted the modifications proposed by \cite{Fukushima2020}.
To overcome the zero division in evaluating the arguments of the arctangent function, \cite{Fukushima2020} replaced  $\tan^{-1} \big( \frac{S}{T} \big)$ by 
\begin{equation}
arctan2(S,T) = \begin{cases}
	%\left\{ \begin{aligned}
    atan (S/T) & \:\: \mbox{if} \: \:\:  T \neq 0 \\
    \pi /2 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S > 0 \\
    -\pi /2 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S < 0 \\
    0 & \:\: \mbox{if} \: \:\:  T = 0  \: \: \mbox{and} \: \:S = 0 \\
    %\end{aligned} \right\}.
    \end{cases}
\label{eq:arctan2}  
\end{equation}
Additionally, if the argument of the logarithm is less than $10^{-10}$, the logarithm is replaced by zero; otherwise the logarithm is  calculated regularly

@@

Because we defined the $z-$axis as positive downwards, the positive vertical displacement means a subsidence and the negative vertical displacement means an uplift.
Figure \ref{fig:displacement}  shows the horizontal and vertical displacements  calculated, respectively, with equations \ref{eq:horizontal_displacement} and \ref{eq:u_til_z} by our methodology that uses the closed expressions of the full integrations (equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}) of \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1} $-$ \ref{dzz2} ).

@@

In both cases (Figures \ref{fig:displacement}b and \ref{fig:displacement_Geertsma}b) the vertical displacements due to the entire the disk-shaped reservoir display a subsidence (positive values) above the reservoir and an uplift (negative values) below the reservoir.
We stress that the proposed volume integrations  (equations \ref{dx1} $-$ \ref{dzz2})  allowed  to evaluate the  vertical displacement (Figure \ref{fig:displacement}b) throughout  the entire reservoir including inside and outside the reservoir.
Rather, the  vertical displacement using Geertsma’s method 
(Figure \ref{fig:displacement_Geertsma}b) is only valid outside the reservoir.

@@

\codedataavailability{
The current version of our code is freely distributed under the BSD 3-clause licence and it is available for download at Zenodo: http://doi.org/XXXXXXXX. 
The latest development version of our code can be freely downloaded from a repository on GitHub (https://github.com/pinga-lab/DisReserv). 
Instructions for running  the current version of our code are also provided on the repository.
The code is still being improved and we encourage the user to work with the latest development version. 
The code was developed as an open-source Python language (Python 3.7.6).
The numerical applications were produced in Jupyter Notebook. 
The data of the pore pressure distribution simulating a realistic reservoir 
$(realistic-model.pickle)$ are available  in the  above-mentioned repositories.
} %% use this section when having data sets and software code available

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement.png}
\caption{Reservoir under uniform depletion: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dzz2})}
\label{fig:displacement}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_z_levels.png}
\caption{Reservoir under uniform depletion: (a) Horizontal x-component displacement and (b) vertical displacement by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}-\ref{dzz2}).
These displacements are calculated along the x-axis, at $y = 0$ m and $z$ located at the depths of:  seafloor ($z = 0$ m), reservoir top ($z = 750$ m), reservoir center ($z = 800$ m) and reservoir bottom ($z = 850$ m).}
\label{fig:displacement_z_levels}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress.png}
\caption{Reservoir under uniform depletion: (a) $x-$, (b) $y-$, and (c) $z-$components of the stress at the free surface.}
\label{fig:Null_stress}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}--\ref{dzz2})}
\label{fig:displacement_non_uniform_depletion}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_z_levels_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) Horizontal x-component displacement and (b) vertical displacement by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}--\ref{dzz2}).
These displacements are calculated along the x-axis, at $y = 0$ m and $z$ located at the depths of:  seafloor ($z = 0$ m), reservoir top ($z = 750$ m), reservoir center ($z = 800$ m) and reservoir bottom ($z = 850$ m).}
\label{fig:displacement_z_levels_non_uniform_depletion}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress_non_uniform_depletion.png}
\caption{Reservoir under non uniform depletion: (a) $x-$, (b) $y-$, and (c) $z-$components of the stress at the free surface.}
\label{fig:Null_stress_non_uniform_depletion}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=8.3cm]{Fig/Figure_Displacement_complex_reservoir.png}
\caption{Reservoir with arbitrary geometry and under arbitrary pressure changes: (a) Horizontal displacement (equation \ref{eq:horizontal_displacement}) and (b) vertical displacement (equation \ref{eq:u_til_z}) by our methodology that uses the closed expressions of the volume integrations given by \cite{Nagyetal2000} and \cite{Nagyetal2002} (equations \ref{dx1}--\ref{dzz2})}
\label{fig:displacement_complex_reservoir}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=12cm]{Fig/Figure_Null_stress_complex_reservoir.png}
\caption{Reservoir with arbitrary geometry and under arbitrary pressure changes: (a) $x-$, (b) $y-$, and (c) $z-$components of the stress at the free surface.}
\label{fig:Null_stress_complex_reservoir}
\end{figure}
